### 2021-03-26 chapters/2-background_and_related_work.tex
finished

This chapter covers three important topics that the following chapters rely on. Identity Management Provisioning as approach to solve problems of distributed partial identities is described along with possible services and systems. The most important regulations of the Online Access Law as well as its basic use case and resulting system architecture of a federal state are described. Messaging is presented as a way of designing technological integration architectures. Multiple messaging patterns as well as their visual notation are explained.

>>>>

This chapter covers three important topics that the following chapters rely on. Identity Management Provisioning as an approach to solve problems of distributed partial identities is described along with possible services and systems. The most important regulations of the Online Access Law as well as its basic use case and the resulting system architecture of a federal state are described. Messaging is presented as a way of designing technological integration architectures. Multiple messaging patterns as well as their visual notation are explained.

### 2021-03-26 chapters/sections/1-context.tex
finished

Today, many services are accessible through the internet. Especially during the COVID-19 pandemic, stores had to switch to online trade due to lockdowns. Online Service Providers (SP) often require customers to create user profiles in order to access their services. Management of personal information as well as interaction and communication between Service Provider and customer are often only possible through a web page or e-mail. It is time consuming to maintain even a small amount of user profiles: To change personal information, like an e-mail address, the web page of each Service Provider has to be accessed the correct credentials have to be submitted and the information has to be manually filled in each time.

In some cases, user profiles are required by Service Providers even if they are not necessary. Especially in case of small stores, running their individual online shops, user profiles are not necessary for customers to place an order. Personal information can be shared temporarily until processing of the order is completed and the package delivered. Some online shops allow users to place orders by temporarily sharing personal information, which however, limits interaction and communication possibilities to insecure e-mail exchange.

A currently relevant topic with an interesting approach for identity management is the German Online Access Law (Online Zugangs Gesetz - OZG). It requires German federal republic, each federal state and commune to provide administration portals for digital access to administrative services. Federal republic and each member state operate their individual user profile systems accessible through their administration portals.

The Online Access Law therefore constructs a system landscape which demonstrates the scenario of distributed partial identities on a small scale while maintaining the same problems as the private sector: Users have to create, maintain and login into multiple user profiles to access administrative services distributed across administration portals. To solve this problem, a solution for making OZG user profiles "interoperable" is in development. However, interoperable user profiles can not be accessed by the private sector.

Identity Management Provisioning (IMP) is an approach to identity management which aims to solve problems of distributed partial identities by enabling users to bring their online identities with them, when accessing systems of Service providers. It provides identity management as a service to users and Service Providers by enabling users to create and manage a single digital identity and use it to share personal information and interact with multiple Service Providers through "IMP relationships" and "IMP messages". Through IMP relationships, users and Service Providers have a tool to securely share personal information which, depending on the use case can be utilized in different ways. It can for example be used to create and manage a persistent SP user profile or to temporarily share personal information for a business process of a SP. IMP messages enable user and Service Provider to securely exchange structured data and depending on the client application which receives and displays the data, design rich user interactions. This approach to identity management could be an alternative to user profiles in the private sector and in context of the OZG.

Identity management systems do not stand alone but have to be accessible by various other systems in a system architecture to for example access personal information and perform authentication. They also play an important role in the design of business processes. Therefore, for Service Providers to utilize IMP, integration into existing business processes and existing system architectures is necessary.

>>>>

Today, many services are accessible through the internet. Especially during the COVID-19 pandemic, stores had to switch to online trade due to lockdowns. Even after lockdowns, customers might continue to favour e-commerce \cite{9076858}. Online Service Providers (SP) often require customers to create user profiles to access their services. Management of personal information as well as interaction and communication between Service Providers and customers are often only possible through a web page or e-mail. It is time consuming to maintain even a small number of user profiles: To change personal information, for example, an e-mail address, the web page of each Service Provider has to be accessed, the correct credentials have to be submitted, and the information has to be manually filled in each time.

In some cases, user profiles are required by Service Providers even if they are not necessary. Especially in the case of small stores, running their individual online shops, user profiles are not necessary for customers to place an order. Personal information can be shared temporarily until processing of the order is completed and the package is delivered. Some online shops allow users to place orders by temporarily sharing personal information, however this limits interaction and communication possibilities to insecure e-mail exchanges.

A currently relevant topic with an interesting approach for identity management is the German Online Access Law (Online Zugangs Gesetz - OZG). It requires the German federal republic, each federal state and each commune to provide administration portals for digital access to administrative services \cite{ozg:general}. The federal republic and each member state operate their individual user profile systems accessible through their administration portals \cite{ozg:user_profiles_general}.

The Online Access Law therefore constructs a system landscape that demonstrates the scenario of distributed partial identities on a small scale while maintaining the same problems as the private sector: Users have to create, maintain, and login into multiple user profiles to access administrative services distributed across administration portals. To solve this problem, a new solution for making OZG user profiles "interoperable" is in development \cite{ozg:user_profiles_general}. However, interoperable user profiles can not be accessed by the private sector.

Identity Management Provisioning (IMP) is an approach to identity management which aims to solve the problems of distributed partial identities by enabling users to bring their online identities with them when accessing the systems of Service Providers. It provides identity management as a service to users and Service Providers by enabling users to create and manage a single digital identity and use it to share personal information and interact with multiple Service Providers through "IMP relationships" and "IMP messages". With IMP relationships, users and Service Providers have a tool to securely share personal information which, depending on the use case, can be utilized in different ways. It can, for example, be used to create and manage a persistent SP user profile or to temporarily share personal information for the business process of a SP. IMP messages enable the user and Service Provider to securely exchange structured data and depending on the client application that receives and displays the data, design rich user interactions. This approach to identity management could be an alternative to user profiles in the private sector and in the context of the OZG.

Identity management systems do not stand alone but have to be accessible by various other systems in a system architecture to, for example, access personal information and perform authentication. They also play an important role in the design of business processes. Therefore, for Service Providers to utilize IMP, integration into existing business processes and existing system architectures is necessary.

### 2021-03-26 chapters/sections/10-imp_solution_integration.tex
finished

The IMP solution presented in this section describes the utilization of IMP relationships with the purpose of creating and maintaining user profiles through the IMP client. Message types for attribute synchronization, authentication and communication are defined. To not disrupt the operation of the existing system architecture, the IMP solution is designed to not replace user profiles but to provide an additional way of accessing them through IMP. To later reduce technological integration efforts, the IMP solution is designed to only require interaction with systems of the portal domain.

\subsection{Relationship Utilization}
The option of using an existing IMP identity for creation of an OZG user profile is added to the account creation web page of the administration portal. An IMP relationship is used for the account creation process. The web page renders the ID of the following relationship template as a QR-code:

@@

Users visiting the account creation web page of the administration portal can use the IMP client to scan the QR code. The IMP client presents the corresponding relationship template to the user and automatically fills in attributes that the administration portal requests from the user as part of the relationship. It is assumed that the IMP system has systems in place for identifying users and validating personal information of corresponding IMP identities. As a result of that, Service Providers can rely on the validity of shared attributes.

@@

\subsection{Message Utilization}
As part of an established relationship, Service Providers and users can securely exchange structured data in form of asynchronous messages. Security in this case is defined as that a messages can not be accessed or modified by a third party and are authorized by the corresponding user. Based on this secure data exchange capabilities, Service Providers can define various types of messages, each with different purposes and effects on the IMP client. This section contains various message types useful for managing the user profile created as a result of an established relationship.

@@

On the login web page of the administration portal, an option is added to login to a user profile using IMP. After entering and submitting the username specified during the profile creation process, administration portal and IMP client exchange authentication messages. Through the IMP client, the user will be requested to approve a login attempt along with information about the corresponding relationship, IP address, URL and more.

@@

The IMP client is used for management of personal information of the user profile. Through the IMP client, the user can send requests for adding, updating or deleting shared attributes. Administration portal and IMP client exchange attribute change request and reply messages. The requests can either be accepted or rejected by the Service Provider. Accepting the request will result in according modification of the user profile connected to the relationship.

For simplification purposes, the web page for the management of personal information on the administration portal is expected to be read-only. However, it would also be possible for the portal to send attribute change requests to the IMP client based on the modifications the user performed on the web page.

@@

User profiles are known to the existing system architecture and are an often used identity management solution. Many Service Providers, especially in the private sector will want to keep using a profile based identity management approach. The IMP solution presents an opportunity of leveraging existing user profile systems while simplifying identity management through IMP. In addition to that, the IMP solution is designed, so that it can almost effortlessly be turned on and off, giving Service Providers the freedom of testing the solution. Service Providers can initially add the IMP solution as an alternative to the conventional user profile identity management solution and after a certain period of time switch to only using the IMP solution.

Mapping user profiles to IMP relationships has many of the usability advantages described in section 3.2. The onboarding process is simplified. Users don't have to manually fill in a form for the creation of a user profile but are supported by the IMP client. As part of established relationships, Service Provider and users are able to directly exchange messages with structured content. In addition to interaction capabilities of the IMP client, this enables a variety of interaction scenarios not possible through for example common e-mail clients. In this case, scenarios for attribute changes, user profile authentication and communication are made possible.

This approach also increases data protection by providing the user with information on exactly which personal information is shared and how it is being processed as part of the profile creation process. This helps users in making an informed decision on whether to create the user profile. At any time after the onboarding process, users can view the relationship in the IMP client and check which Service Providers process which personal information, why and how. This enables users to keep an overview of the distribution of their personal information.

Security is also increased as a result of this approach. Through established relationships, Service Provider and users have the ability to not only directly exchange messages but also to do this securely. As a result of that, authentication through the IMP client can replace password based authentication methods. This eliminates the risk of weak or stolen passwords.

In case of the OZG, this IMP solution can eventually replace the need for interoperable user profiles. In case each administration portal enables onboarding as described by the IMP solution, regarding usability, the existence of a separation of user profiles can be transparent to the user but regarding data protection, the user profiles are strongly separated:

After an effortless onboarding process on multiple administration portals, users are able to manage each profile through the IMP client. This enables users to break free access from all user profiles giving the illusion of an interoperable user profile. Logging in to a user profile using the IMP client eliminates the need of remembering passwords, furthering the illusion of interoperable user profiles. Finally, if for each user profile, the same profile name is selected, the user wont be able to distinguish user profiles most of the time.

However in cases where the user requires distinction between user profiles in regard to data protection, the IMP client enables the user to separately view personal information shared with each portal and information about how each portal processes it.

@@

The IMP solution has also several problems regarding usability and data protection. Most importantly, as a result of integrating the IMP system through user profiles, each administration portal is still required to access personal information. In addition to that, application and management of administrative services remains accessible only through the administration portal and form server which presents a discontinuity in accessing OZG services.

>>>>

The IMP solution presented in this section describes the utilization of IMP relationships with the purpose of creating and maintaining OZG user profiles through the IMP client. Message types for attribute synchronization, authentication, and communication are defined. To not disrupt the operation of the existing system architecture, the IMP solution is designed to not replace user profiles but to provide an additional way of accessing them through IMP. To later reduce technological integration efforts, the IMP solution is designed to only require interaction with systems of the portal domain.

\subsection{Relationship Utilization}
The option of using an existing IMP identity for the creation of an OZG user profile is added to the account creation web page of the administration portal. An IMP relationship is used for the account creation process. The web page renders the ID of the following relationship template as a QR-code:

@@

Users visiting the account creation web page of the administration portal can use the IMP client to scan the QR code. The IMP client presents the corresponding relationship template to the user and automatically fills in attributes that the administration portal requests from the user as part of the relationship. It is assumed that the IMP system has systems in place for identifying users and validating the personal information of corresponding IMP identities. As a result of that, Service Providers can rely on the validity of shared attributes.

@@

\subsection{Message Utilization}
As part of an established relationship, Service Providers and users can securely exchange structured data in the form of asynchronous messages. Security in this case is defined as that a message can not be accessed or modified by a third party and are authorized by the corresponding user. Based on this secure data exchange capability, Service Providers can define various types of messages, each with different purposes and effects on the IMP client. This section contains various message types useful for managing the user profile created as a result of an established relationship.

@@

On the login web page of the administration portal, an option is added to login to a user profile using IMP. After entering and submitting the username specified during the profile creation process, the administration portal and IMP client exchange authentication messages. Through the IMP client, the user will be requested to approve a login attempt along with information about the corresponding relationship, IP address, URL, and more.

@@

The IMP client is used for the management of personal information of the user profile. Through the IMP client, the user can send requests for adding, updating, or deleting shared attributes. Administration portal and IMP client exchange attribute change request and reply messages. The requests can either be accepted or rejected by the Service Provider. Accepting the request will result in according modification of the user profile connected to the relationship.

For simplification purposes, the web page for the management of personal information on the administration portal is expected to be read only. However, it would also be possible for the portal to send attribute change requests to the IMP client based on the modifications the user performed on the web page.

@@

User profiles are known to the existing system architecture and are often used as an identity management solution. Many Service Providers, especially in the private sector, will want to keep using a profile-based identity management approach. The IMP solution presents an opportunity of leveraging existing user profile systems while simplifying identity management through IMP. In addition to that, the IMP solution is designed so that it can almost effortlessly be turned on and off, giving Service Providers the freedom of testing the solution. Service Providers can initially add the IMP solution as an alternative to the conventional user profile identity management solution and after a certain period of time switch to only using the IMP solution.

Mapping user profiles to IMP relationships has many of the usability advantages described in section 3.2. The onboarding process is simplified. Users do not have to manually fill in a form for the creation of a user profile but are supported by the IMP client. As part of established relationships, Service Providers and users are able to directly exchange messages with structured content. In addition to the interaction capabilities of the IMP client, this enables a variety of interaction scenarios not possible through, for example, common e-mail clients. In this case, scenarios for attribute changes, user profile authentication, and communication are made possible.

This approach also increases data protection by providing the user with information on exactly which personal information is shared and how it is being processed as part of the profile creation process. This helps users in making an informed decision on whether to create the user profile. At any time after the onboarding process, users can view the relationship with the IMP client and check which Service Providers process which personal information, why, and how. This enables users to keep an overview of the distribution of their personal information.

Security is also increased as a result of this approach. Through established relationships, Service Providers and users have the ability to not only directly exchange messages but also to do this securely. As a result of that, authentication through the IMP client can replace password-based authentication methods. This eliminates the risk of weak or stolen passwords.

In the case of the OZG, this IMP solution can eventually replace the need for interoperable user profiles. In case each administration portal enables onboarding as described by the IMP solution, regarding usability, the existence of a separation of user profiles can be transparent to the user, but regarding data protection, the user profiles are strongly separated:

After an effortless onboarding process on multiple administration portals, users are able to manage each profile through the IMP client. Logging in to a user profile using the IMP client eliminates the need of remembering passwords, giving the illusion of interoperable user profiles. Finally, if for each user profile, the same profile name is selected, the user wont be able to distinguish OZG user profiles most of the time.

However, in cases where the user requires a distinction between OZG user profiles regarding data protection, the IMP client enables the user to separately view the personal information shared with each portal and information about how each portal processes it.

@@

The IMP solution has also several problems regarding usability and data protection. Most importantly, as a result of integrating the IMP system through user profiles, each administration portal is still required to access personal information. Therefore, the data protection issues described in section \ref{chapter:user_profile_identity_management} remain. In addition to that, application and management of administrative services remains accessible only through the administration portal and form server, which presents a discontinuity in accessing OZG services.
